---
title: "Full Pipeline"
format: html
editor: visual
---

# Load Packages

```{r}
set.seed(99)
library(dplyr)
library(ggplot2)
library(DESeq2)
library(ggpubr)
library(Seurat)
library(RColorBrewer)
library(ggrepel)
library(harmony)
library(gridExtra)
library(rstudioapi)
library(patchwork)
library(S4Vectors)
library(tidyverse)
library(sctransform)
library(decontX)
library(glmGamPoi)
library(EnhancedVolcano)
```

Set working directory and read in Seurat objects(.rds).

```{r}
pnt <- "C006"
wkdir <- getwd()
rds <- readRDS( paste0(wkdir,'/Projects/C6-exact-cutadapt-noIntron_Seurat.rds'))

# Run after previously clustered
#rds <- readRDS(paste0(savePath, cart, ".RDS"))
```

Separate WTA from AbSeq (depreciated)

Create assay for each and delete original RNA assay\
Abseq data is isolated by grepping rows that end with "pAbO".

adt \<- rds\@assays$RNA$counts\[grep("pAbO", rownames(rds)),\] #print(head(abseq)) rna \<- rds\@assays$RNA$counts\[-(grep("pAbO", rownames(rds))),\] rds\[\["ADT"\]\] \<- CreateAssayObject(adt) \## Warning: Feature names cannot have '\_', replacing with '-' DefaultAssay(rds) = "RNA" rm(adt) rm(rna) gc()

[Abseq panel](https://docs.google.com/spreadsheets/d/1ZKzQnN4Y-RBqatd2ftfxB4pQZeHfals3cGp6iybdvqk/edit#gid=427554242)

### Save Feature Names

Closer look at newly created assays. Store Abseq names and WTA genes for later feature call\
BD Rhapsody Enhanced Cell Capture Beads have

```{r}
# for C29 should be 31 Abseq markers
abseq <- rownames(rds[["ADT"]])
all.genes <- rownames(rds[["RNA"]])

#rownames(obj[["RNA"]])
#colnames(obj[["RNA"]])
dim(rds[["RNA"]])
```

## Quality Check

Threshold for low reads and high mitochondrial read content (dead cells)

EoE_Hal original dimensions: col = 38659 row = 140856

TODO: add a lower bound for nfeat/ncount

```{r}
# add useful metadata to the Seurat object 
# in this case the percent of mitochondrial reads 
# looks for genes with the mitochondrial tag, always possible to miss some 
rds[["percent.mt"]] <- PercentageFeatureSet(rds, pattern = "^MT-") 
features = c("nFeature_RNA", "nCount_RNA", "percent.mt") 
dim(rds) 
VlnPlot(rds, features = features, pt.size = 0.01, ncol = 3) & labs(x="")
```

Select QC thresholds

```{r}
rds <- subset(rds, subset = nFeature_RNA > 300 & nFeature_RNA < 4000 & nCount_RNA > 150 & nCount_RNA < 30000 & percent.mt < 30)
dim(rds)
VlnPlot(rds, features = features, pt.size = 0.01, ncol = 3) & labs(x="")+ ggtitle(ptn)
```

## Normalize WTA & Abseq \> PCA

Seurat 5 default norm: SC transform

```{r}
# Normalize across cells for WTA
# TODO updated transform
DefaultAssay(rds) <- 'RNA'
rds <- NormalizeData(rds) %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()

DefaultAssay(rds) <- 'ADT'
# we will use all ADT features for dimensional reduction
# we set a dimensional reduction name to avoid overwriting the 
VariableFeatures(rds) <- rownames(rds[["ADT"]])
rds <- NormalizeData(rds, normalization.method = 'CLR', margin = 2) %>% 
  ScaleData() %>% RunPCA(reduction.name = 'apca')
DefaultAssay(rds) <- 'RNA'

#rds <- SCTransform(rds, verbose = FALSE) %>% RunPCA()
rds <- FindMultiModalNeighbors(
  rds, reduction.list = list("pca", "apca"), 
  dims.list = list(1:30, 1:18), modality.weight.name = "RNA.weight"
) 
```

### Variable Genes

```{r}
# Identify the 10 most highly variable genes
feat <- 50
top <- head(VariableFeatures(rds), feat)
bot <- tail(VariableFeatures(rds), feat)

rds <- FindVariableFeatures(rds, selection.method = "vst", nfeatures = 2000)

# plot variable features with and without labels
varPlot <- VariableFeaturePlot(rds)

varPlot <- LabelPoints(varPlot, points = c(top), repel = TRUE) + 
ggtitle(paste("Overall Variable Features:"))

# Write most variable genes to table if you want
write.csv(top, paste0(savePath, paste0(cart,"_top_all_expression.csv")))
write.csv(bot,paste0(savePath, paste0(cart,"bot_all_expression.csv")))

varPlot
```

## Clustering

#### UMAP

We use first 10-20ish PCs to cluster, wherever elbow plot drops off\
Try to capture as much variance as possible\
Umap Louvain algorithm: higher res = higher \# of clusters

#### WNN

```{r}
# WNN CLUSTERING
DefaultAssay(rds) <- "RNA"
dim <- 16
rds <- FindMultiModalNeighbors(
  rds, reduction.list = list("pca", "apca"), 
  dims.list = list(1:30, 1:18), modality.weight.name = "RNA.weight"
)
rds <- RunUMAP(rds, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
rds <- FindClusters(rds, graph.name = "wsnn", algorithm = 3, resolution = 0.8, verbose = FALSE)

```

FindCluster algorithms:

1 - Louvain algorithm

2 - Louvain w/ multilevel refinement

3 - SLM alg

4 - Leiden alg (requires the leidenalg python)

```{r}
res <- 0.8
rds <- RunUMAP(rds, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
rds <- FindClusters(rds, graph.name = "wsnn", algorithm = 3, resolution = 0.8, verbose = FALSE)

rdsClust <- FindClusters(rds, graph.name = "wsnn", algorithm = 1, resolution = res, verbose = FALSE)

DimPlot(rds, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend() + plot_annotation(title = paste("WNN UMAP res: Louvian:", res, ":", cart))
gc()
```

```{r}
rdsClust <- FindClusters(rds, graph.name = "wsnn", algorithm = 2, resolution = res, verbose = FALSE)

DimPlot(rdsClust, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend() + plot_annotation(title = paste("WNN UMAP res: Louvian w/multilevel refinement:", res, ":", cart))
gc()
```

```{r}
rdsClust <- FindClusters(rds, graph.name = "wsnn", algorithm = 3, resolution = res, verbose = FALSE)

DimPlot(rdsClust, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend() + plot_annotation(title = paste("WNN UMAP res: SLM alg", res, ":", cart))
gc()
```

```{r}
rdsClust <- FindClusters(rds, graph.name = "wsnn", algorithm = 4, resolution = res, verbose = FALSE)

DimPlot(rdsClust, reduction = 'wnn.umap', label = TRUE, repel = TRUE, label.size = 2.5) + NoLegend() + plot_annotation(title = paste("WNN UMAP res: Leiden alg", res, ":", cart))
gc()
```

###### WNN Dotplot

```{r}
cart<-'C006'
DefaultAssay(rds) <- "ADT"
DotPlot(rds, features = rownames(GetAssayData(rds, assay = 'ADT', slot = "counts")), cols = "RdBu", col.min = -1, dot.scale = 8, cluster.idents = TRUE) + 
  coord_flip() + labs(title = paste("AbSeq Dotplot:", cart))
# reset the default assay back to RNA!!
DefaultAssay(rds) <- "RNA"
```

Feature plots

WTA umap data -\> overlay Abseq expression data

```{r}
#abseq = rownames(rds@assays$ABSEQ)
DefaultAssay(rds) <- "ADT"
# for each of the 31 abseq markers, make a feature plot
FeaturePlot(rds, features = abseq[0:5], ncol = 5)
FeaturePlot(rds, features = abseq[6:10], ncol = 5)
FeaturePlot(rds, features = abseq[11:15], ncol = 5)
FeaturePlot(rds, features = abseq[16:20], ncol = 5)
FeaturePlot(rds, features = abseq[21:25], ncol = 5)
FeaturePlot(rds, features = abseq[26:30], ncol = 5)
FeaturePlot(rds, features = abseq[31])
DefaultAssay(rds) <- "RNA"
```

Ridge Plots

```{r}
RidgePlot(rds, assay = "ADT", features = abseq[0:8], ncol = 8)
RidgePlot(rds, assay = "ABSEQ", features = abseq[9:16], ncol = 8)
RidgePlot(rds, assay = "ABSEQ", features = abseq[17:24], ncol = 8)
RidgePlot(rds, assay = "ABSEQ", features = abseq[25:31], ncol = 8)
```

Violin plots: Show Abseq expression for each cluster

```{r}
# chunk "21": violin plots for first 4 Abseq features
VlnPlot(object = obj,features = abseq[0:4], pt.size = 0, assay = "Abseq", ncol = 2) & theme(plot.title = element_text(size = 10), axis.text.x = element_text(angle = 0, hjust = 0.5), axis.title.x = element_blank())
```

Pull subset of CD119+ cells

```{r}
eos <- rds$assay@ABSEQ
```

## Cluster Annotation

```{r}
#obj <- RunAzimuth(obj, reference = "pbmcref")

###############  ANNOTATING CLUSTERS ############### 
annotations <- rbind(c("0", "Neutrophil progenitor"),
                     c("1", "Noise"),
                     c("2", "Neutrophil progenitor"),
                     c("3", "Eosinophil"),
                     c("4", "T Cell"),
                     c("5", "Monocyte"),
                     c("6", "T Cell"),
                     c("7", "T Cell"),
                     c("8", "Noise"),
                     c("9", "Neutrophil"),
                     c("10", "Neutrophil"),
                     c("11", "T Cell"),
                     c("12", "Natural Killer"),
                     c("13", "Platelet"),
                     c("14", "B cell"),
                     c("15", "Monocyte"),
                     c("16", "Activated B cell"),
                     c("17", "Basophil"),
                     c("18", "HPSC"))
colnames(annotations) <- c("Cluster", "CellType")
annotations <- data.frame(annotations)

# prepare the annotation information
annotations <- setNames(annotations$CellType, annotations$Cluster)

# relabel the Seurat clusters
# Idents(obj) <- "seurat_clusters"
rds <- RenameIdents(rds, annotations)

# alphabetize the cell types
Idents(rds) <- factor(Idents(object = rds), levels = sort(levels(rds)))

# useful metadata (e.g. if you want to have multiple annotation sets)
rds[["annotated_clusters"]] <- Idents(object = rds)

# info about the clusters
obj_annotations %>%
  group_by(CellType) %>%
  transmute(Clusters = paste0(Cluster, collapse = ", ")) %>%
  distinct() %>%
  arrange(CellType)


UMAPPlot(rds, label = T) + plot_annotation(title = paste("Annotated UMAP:", cart)) + theme(legend.position="none")
## UMAP Cells/cluster
write.csv(t(table(Idents(rds))), file=paste0(savePath, paste0(cart, "_clusters.csv")))

DotPlot(rds, features = rownames(GetAssayData(object = rds@assays$ADT, slot = "counts")), 
        cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE) + coord_flip() + RotatedAxis() + plot_annotation(title = paste("AbSeq Dotplot:", cart))

# metadata file for MCIA
meta <- data.frame("CellType" = Idents(rds))
```

# SAVE YOUR WORK

```{r}
setwd("C:/Users/Innoscan/Documents/Innoscan_Rhapsody")# Work PC
#setwd("~/Research/R/Home Rhapsody")                  # Home PC

saveRDS(rds, file= paste0(savePath, cart, ".RDS"))
#sessionInfo()
```
