---
title: "Rhapsody Pipeline"
editor: visual
date: "Compiled: `r format(Sys.time(), '%d %B %Y')`"
format: pdf
---

## CiteSeq

ST 11: 302 EoE post dupi

ST 12: 313 EoE control

7B cell call:

```{r}
# chunk 1: load libraries
library(GEOquery)
library(dplyr)
library(ggplot2)
library(DESeq2)
library(ggpubr)
library(Seurat)

#library(pheatmap)
library(RColorBrewer)
#library(ggdendro)
#library(edgeR)
library(ggrepel)
#library(rgl)
library(gplots) # heatmap.2
#library(factoextra)
#library(dendextend)
library(plotly)
```

Maybe limit R virtual memory access? HAVE NOT TRIED RUNNING

```{r}
#memory.limit(size=2500)
```

Set working directory and read RDS.

```{r}
cart <- "C029"
setwd("C:/Users/Innoscan/Documents/Innoscan_Rhapsody")
setwd("~/Research/R/Home Rhapsody")
savePath <- paste0("RDS_files/",cart, "/")
rds <- readRDS(file = paste0(savePath, "C36-exact_Seurat.rds"))

# Run after previously clustered
# rds <- readRDS(paste0(savePath, "clustered.RDS"))
```

TODO Separate by SMK

```{r}

DefaultAssay(rds) <- 'ADT'
# we will use all ADT features for dimensional reduction
# we set a dimensional reduction name to avoid overwriting the 
VariableFeatures(rds) <- rownames(rds[["ADT"]])
rds <- NormalizeData(rds, normalization.method = 'CLR', margin = 2) %>% 
  ScaleData() %>% RunPCA(reduction.name = 'apca')
# Identify multimodal neighbors. These will be stored in the neighbors slot, 
# and can be accessed using dress[['weighted.nn']]
# The WNN graph can be accessed at dress[["wknn"]], 
# and the SNN graph used for clustering at dress[["wsnn"]]
# Cell-specific modality weights can be accessed at dress$RNA.weight
rds <- FindMultiModalNeighbors(
  rds, reduction.list = list("pca", "apca"), 
  dims.list = list(1:30, 1:18), modality.weight.name = "RNA.weight"
)
rds <- RunUMAP(rds, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
rds <- FindClusters(rds, graph.name = "wsnn", algorithm = 3, resolution = 0.6, verbose = FALSE)
```

Separate WTA from AbSeq data and create assay for each. (RNA = WTA, AbSeq = ADT)

Abseq data is isolated by grepping rows that end with "pAbO".

[C29 Abseq panel](https://docs.google.com/spreadsheets/d/1ZKzQnN4Y-RBqatd2ftfxB4pQZeHfals3cGp6iybdvqk/edit#gid=427554242)

```{r}
adt <- obj@assays$RNA$counts[grep("pAbO", rownames(obj)),]
rna <- obj@assays$RNA$counts[-(grep("pAbO", rownames(obj))),]
obj[["Abseq"]] <- CreateAssayObject(adt)
## Warning: Feature names cannot have underscores ('_'), replacing with dashes ('-')
# See assays, set default to WTA
rds[["WTA"]] <- CreateAssayObject(rna)
DefaultAssay(rds) = "WTA"
rds[['RNA']] <- NULL
Assays(rds)
```

Closer look at newly created assays

BD Rhapsody Enhanced Cell Capture Beads have

```{r}
# for C29 should be 31 Abseq markers
rownames(rds[["ABSEQ"]])

#rownames(obj[["WTA"]])
#colnames(obj[["WTA"]])
dim(rds[["WTA"]])

# add useful metadata to the Seurat object
# in this case the percent of mitochondrial reads
# looks for genes with the mitochondrial tag, always possible to miss some
rds[["percent.mt"]] <- PercentageFeatureSet(rds, pattern = "^MT-")
```

Quality Check: threshold for low reads and high mitochondrial read content (dead cells)

TODO: add a lower bound for nfeat/ncount

```{r}
# chunk "7": create violin plots for QC
# try + vs & 
features = c("nFeature_RNA", "nCount_RNA", "percent.mt")
VlnPlot(rds, features = features,
        pt.size = 0.01, ncol = 3) & labs(x="")
rds <- subset(x = rds, subset = nFeature_RNA > 200 & nFeature_RNA < 6500 & percent.mt < 15)

# Violin plot after QC
VlnPlot(rds, features) & labs(x="")

# Save WTA genes as variable for later
genes <- rownames(rds[["WTA"]])

dim(rds)
```

Normalize WTA and AbSeq separately

Seurat 5 default norm: SC transform

```{r}
# Normalize across cells for WTA
# TODO updated transform
rds <- NormalizeData(rds, normalization.method = "LogNormalize",
                     scale.factor = 10000, assay = "WTA")
# log1p(exp(1)-1)

## Normalize across cells for Abseq
rds <- NormalizeData(rds, normalization.method = "CLR",
                     margin = 2, assay = "ABSEQ")  
#margin=1 is features margin=2 is across cells

# highly variable features
# does it only on the RNA because "Active Assay" is RNA
rds <- FindVariableFeatures(rds, selection.method = "vst",
                            nfeatures = 2000)
rds <- RunPCA(rds)


VariableFeatures(rds) <- rownames(rds[["ABSEQ"]])
rds <- NormalizeData(rds) %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()
DefaultAssay(rds) <- 'ABSEQ'
VariableFeatures(rds) <- rownames(rds[["ADT"]])
rds <- NormalizeData(rds, normalization.method = 'CLR', margin = 2) %>% 
  ScaleData() %>% RunPCA(reduction.name = 'apca')
DefaultAssay(rds) <- 'WTA'

# scaling so the average expression is 0 and the variance is 1
# only for RNA component 
rds <- ScaleData(rds, features = rownames(rds))
DefaultAssay(rds) <- "ABSEQ"
rds <- ScaleData(rds, features = rownames(rds))
DefaultAssay(rds) <- "WTA"
#obj@meta.data
```

Closer look at variable genes

```{r}
# Identify the 10 most highly variable genes
feat <- 50
top <- head(VariableFeatures(rds), feat)
bot <- tail(VariableFeatures(rds), feat)

rds <- FindVariableFeatures(rds, selection.method = "vst", nfeatures = 2000)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(rds)
plot2 <- LabelPoints(plot = plot1, points = c(top), repel = TRUE) + title(paste(cart, "Overall DEG"))

all.genes.rds <- rownames(rds)

plot2
```

PCA

Default assay is WTA so PCA is run on just WTA

```{r}
#dim(obj)
PCAPlot(rds)

# chunk "11": plot the variance explained by addition of each PCA component 
# to determine the no of PCs to use for downstream analysis
ElbowPlot(rds, ndims = 30)
head(VariableFeatures(rds), 20)
# these can be plotted if desired
#LabelPoints(plot = VariableFeaturePlot(object = obj),
#            points = head(VariableFeatures(object = obj), 20),
#            repel = TRUE, xnudge = 0, ynudge = 0)

# Heatmap view of desired PCs, here we look at first 5
DimHeatmap(obj, dims = 1:5,cells = 500, balanced = TRUE)

# see genes for each PC
VizDimLoadings(obj, dims = 1:2, reduction = 'pca')
```

Attempting 3D PCA for funsies using plotly

```{r}
c29PCA <- Embeddings(obj, "pca")
fig <- plot_ly(c29PCA, x = ~PC1, y = ~PC2, z = ~PC3, color = ~iris$Species, colors = c('#636EFA','#EF553B','#00CC96') )
```

Clustering for UMAP/tSNE/WNN

We use first 20 PCs to cluster, wherever elbow plot drops off

Umap Louvain algorithm: higher res = higher \# of clusters

```{r}
# Computing shared nearest neighbor graph
rds <- FindNeighbors(rds, reduction = "pca", dims = 1:20)

# clustering (adjust dimensions and resolutions as desired)
# default is Louvain clustering - check how it works
# https://www.rdocumentation.org/packages/Seurat/versions/1.4.0/topics/FindClusters
rds <- FindClusters(rds, resolution = 0.5)

rds <- RunUMAP(rds, reduction = "pca", dims = 1:20)

plot_seurat_clusters <- UMAPPlot(rds, label = TRUE, label.size = 5) +
                          labs(title = paste("Initial UMAP:", cart)) +
                          theme(plot.title = element_text(hjust = 0.5),
                                axis.text = element_blank(),
                                axis.ticks = element_blank()) + 
                          theme(legend.position="none")
plot_seurat_clusters
```

Dotplot using Abseq (skipped WTA Dotplot)

```{r}
# chunk 18: make a dot plot for clusters using ADT data
# you have to change the default assay for the dot plot
DefaultAssay(object = obj) <- "Abseq"

DotPlot(object = obj,
        features = rownames(GetAssayData(object = obj, assay = "Abseq",
                                         slot = "counts")),
        cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE) +
  RotatedAxis()

# reset the default assay back to RNA!!
DefaultAssay(object = obj) <- "RNA"
```

Feature plots:

WTA umap data -\> overlay Abseq expression data

Keep in mind default assay is still WTA

```{r}
abseq = rownames(obj@assays$Abseq)

# for each of the 31 abseq markers, make a feature plot
# show first 6 here to save time
FeaturePlot(obj,features = abseq[0:6], 
            min.cutoff = 0, label = TRUE, label.size = 3,
            ncol = 3, raster = FALSE)

```

Violin plots: Show Abseq expression for each cluster

```{r}
# chunk "21": violin plots for first 4 Abseq features
VlnPlot(object = obj,
        features = abseq[0:4],
        pt.size = 0, assay = "Abseq", ncol = 2) &
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title.x = element_blank())
```

Annotating each cluster

Azimuth test

```{r}
obj <- RunAzimuth(obj, reference = "pbmcref")
```

SAVE YOUR WORK

```{r}
saveRDS(rds, file= saveFileName)
sessionInfo()
```
