---
title: "Single Cell Analysis of Cite-seq data "
author: "Max Mattessich, Joaquin Reyna, Edel Aron, Anna Konstorum"
date: "Compiled: `r format(Sys.time(), '%d %B %Y')`"
format: pdf
editor: visual
---

## Introduction

In this vignette, we will cover some of the possible sources for publicly available single-cell data, how to format it for and process it with MCIA and how to run a basic analysis in order to annotate the data and use the results as metadata for exploring the decomposition results.

**`OUTLINE`**

![](https://drive.google.com/uc?export=view&id=1tmTkjNa8BwQEyyL6Amab1XQFdl6uzCNS)

**`DATA FLOW`**

![](https://drive.google.com/uc?export=view&id=1r0UJXc4-Gp3431NVk7D-60Y1Qmft4F8D){width="1867"}

![](https://drive.google.com/uc?export=view&id=1h4pFQe6gM3k8LI6QsmZ4rFDHo9jimgo6)

```{r}
# chunk 1: load libraries
library(dplyr)
library(ggplot2)
library(ggpubr)
library(Seurat)

#library(pheatmap)
library(RColorBrewer)
#library(ggdendro)
#library(edgeR)
library(ggrepel)
#library(rgl)
library(gplots) # heatmap.2
#library(factoextra)
#library(dendextend)
```

## Data

We will be using the 10x Genomics "pbmc5k-CITEseq" dataset, which was published in May 2019 and processed using Cell Ranger 3.0.2. It contains 5,247 detected cells from PBMCs and 33,538 genes along with 32 cell surface markers. We chose this dataset due to it containing both gene expression (GEX) and cell surface protein (ADT) data as well as being publicly available and from a widely used platform.

The following code details several ways in which to load in this dataset. The original dataset can be found on the \[[10x Genomics Datasets website](https://support.10xgenomics.com/single-cell-gene-expression/datasets/3.0.2/5k_pbmc_protein_v3)\] and an explanation of the file types for this version can be found [here](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/3.0/output/overview). You can download them all to a directory of your choosing with their suggested terminal commands:

```         
```

```{r}
# chunk 2: download files from 10x website
#| echo: false
# recommended location for local data
#dir.create("../data")
path_data <- file.path("..", "data")
dir.create(path_data, showWarnings = FALSE)

getwd()
path_data

if (!file.exists(file.path(path_data,"5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz")))
{
  
  # download the raw feature matrix
  url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_raw_feature_bc_matrix.tar.gz"
dest = file.path(path_data, "5k_pbmc_protein_v3_raw_feature_bc_matrix.tar.gz")
download.file(url,      destfile = dest, method = "auto")

  # download the filtered feature matrix
  # this is after 10x cell ranger filters for barcodes 
  # barcodes that do not correspond to cells - barcode read errors etc
  url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz"
  dest = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz")
  download.file(url, destfile = dest, method = "auto")


# download the metrics summary
  url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_metrics_summary.csv"
  dest = file.path(path_data, "5k_pbmc_protein_v3_metrics_summary.csv")
  download.file(url, destfile = dest, method = "auto")


# download the summary html (optional)
  url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_web_summary.html"
  dest = file.path(path_data, "5k_pbmc_protein_v3_web_summary.html")
  download.file(url, destfile = dest, method = "auto")
}
```

If you would like to view some basic information about the data, you can open the \*web_summary.html\* in your favorite web browser. It will show the estimated number of cells, mean reads per cell, median genes per cell and a variety of other sample metrics. These metrics are also available in the \*metrics_summary.csv\*

You will need the \*filtered_feature_bc_matrix\* directory for the following analysis; it can be extracted with \`tar -xvzf 5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz\` from within the relevant data directory. Or can be done in R as below:

### Extract the data

```{r}
# chunk 3: Extract the data using untar in R
if (!file.exists(file.path(path_data, "tenx_pbmc5k_CITEseq")))
{

tar_fn = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz")
#exdir = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix")
exdir = file.path(path_data, "tenx_pbmc5k_CITEseq")
untar(tar_fn, exdir = exdir, tar = "internal")
}
```

## Seurat Analysis

This section demonstrates how to take in raw data (in this case, the output of Cell Ranger) and go through a popular analysis pipeline to ultimately cluster and annotate the data. Some people prefer to use Cell Ranger's built-in dimension reduction and clustering analysis and to view the results with [Loupe Cell Browser](https://support.10xgenomics.com/single-cell-gene-expression/software/visualization/latest/what-is-loupe-cell-browser).

*Credit to Seurat ([RNA](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) and [multi-modal](https://satijalab.org/seurat/articles/multimodal_vignette.html#setup-a-seurat-object-add-the-rna-and-protein-data)) for the general steps.*

### Load the data in to a Seurat object

```{r}
# chunk 4: load the data (change the file path as needed)
data <- Seurat::Read10X(data.dir = file.path(path_data, "tenx_pbmc5k_CITEseq",                                        "filtered_feature_bc_matrix"),
                        strip.suffix = TRUE) # remove the "-1"s from barcodes
## 10X data contains more than one type and is being returned as a list
## containing matrices of each type.

# set minimum cells and/or features here if you'd like
obj <- Seurat::CreateSeuratObject(counts = data$`Gene Expression`,
                                  project = "pbmc5k_CITEseq")

obj[["ADT"]] <- Seurat::CreateAssayObject(counts = data$`Antibody Capture`)
## Warning: Feature names cannot have underscores ('_'), replacing with dashes ('-')

# check the assays
Seurat::Assays(object = obj)
## "RNA" "ADT"

# list out the CITE-Seq surface protein markers
rownames(obj[["ADT"]])
# list out the RNA gene labels
#rownames(obj[["RNA"]])
#length(colnames(obj))
obj

# add useful metadata to the Seurat object
# in this case the percent of mitochondrial reads
# it looks for genes with the mitochondrial tag, always possible to miss some
obj[["percent.mt"]] <- PercentageFeatureSet(object = obj, pattern = "^MT-")
```

```{r}
# chunk 5: explore what is in each of the three files Seurat wants to read
file=file.path(path_data, "tenx_pbmc5k_CITEseq", "filtered_feature_bc_matrix", "barcodes.tsv.gz")
barcodesData=read.csv(file,header=FALSE)
dim(barcodesData)
head(barcodesData)

file=file.path(path_data, "tenx_pbmc5k_CITEseq", "filtered_feature_bc_matrix", "features.tsv.gz")
featuresData=read.csv(file)
dim(featuresData)
head(featuresData)

file=file.path(path_data, "tenx_pbmc5k_CITEseq", "filtered_feature_bc_matrix", "matrix.mtx.gz")
matrixData=read.table(file,skip=3) # skip=3 is new!! File had headers
dim(matrixData)
head(matrixData)

```

#### Metrics summary

```{r}
# chunk 6: read in the summary table

list.files(path_data)
metrics_summary <- read.csv(
    file = file.path(path_data, "5k_pbmc_protein_v3_metrics_summary.csv")
    )
metrics_summary 
```

#### Gene expression quality control and filtering

```{r}
# chunk 7: create violin plots for QC
# try + vs & 
VlnPlot(object = obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        pt.size = 0.01, ncol = 3) & labs(x="")
dim(obj)
```

Based on the previous plots, a minimum of 200 features and a maximum of 20% mitochondrial seemed like good cutoffs.

```{r}
# chunk 8: remove cells with certain issues - adjust cutoffs as desired
#& nFeature_RNA <4000
obj <- subset(x = obj, subset = nFeature_RNA > 200  & percent.mt < 20)

VlnPlot(object = obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        pt.size = 0.01, ncol = 3) & labs(x="")
dim(obj)
```

#### Seurat data normalization and transformation

```{r}
# chunk 9: Seurat data transformation and normalization 

# standard log normalization for RNA and centered log for ADT
obj <- NormalizeData(object = obj, normalization.method = "LogNormalize",
                     scale.factor = 10000, assay = "RNA")
# log1p(exp(1)-1)

## Normalizing across cells for the ADT data
obj <- NormalizeData(object = obj, normalization.method = "CLR",
                     margin = 2, assay = "ADT")  
#margin=1 is features margin=2 is across cells

# highly variable features
# does it only on the RNA because "Active Assay" is RNA
# can be set by DefaultAssay() function
obj <- FindVariableFeatures(object = obj, selection.method = "vst",
                            nfeatures = 2000)

# scaling so the average expression is 0 and the variance is 1
# only for RNA component 
obj <- ScaleData(object = obj, features = rownames(obj))

#obj@meta.data
```

#### Principal component analysis (PCA) and dimensionality reduction -FUN STUFF!

```{r}
# chunk 10: dimensionality reduction using PCA
obj <- RunPCA(object = obj)
dim(obj)
PCAPlot(obj)
```

```{r}
plot3d(obj, col = type, type = 's', size = 1,
       xlab = 'PC1', ylab = 'PC2', zlab = 'PC3')
```

```{r}
# chunk 11: plot the variance explained by addition of each PCA component 
# to determine the no of PCs to use for downstream analysis
ElbowPlot(object = obj, ndims = 30)
head(VariableFeatures(object = obj), 20)
# these can be plotted if desired
#LabelPoints(plot = VariableFeaturePlot(object = obj),
#            points = head(VariableFeatures(object = obj), 20),
#            repel = TRUE, xnudge = 0, ynudge = 0)

```

```{r}
# chunk 12: Computing shared nearest neighbor graph
obj <- FindNeighbors(object = obj, reduction = "pca", dims = 1:20)
# https://www.rdocumentation.org/packages/Seurat/versions/3.0.1/topics/FindNeighbors
```

```{r}
# chunk 13: clustering (adjust dimensions and resolutions as desired)
# default is Louvain clustering - check how it works
# talk about resolution a bit
obj <- FindClusters(object = obj, resolution = 0.4)
# https://www.rdocumentation.org/packages/Seurat/versions/1.4.0/topics/FindClusters
```

```{r}
# chunk 14: the famous UMAP
obj <- RunUMAP(object = obj, reduction = "pca", dims = 1:20)

# If you want to see a heatmap interpretation of PCs
DimHeatmap(obj, dims = 1:5, balanced = TRUE)

# https://www.rdocumentation.org/packages/Seurat/versions/2.3.4/topics/RunUMAP
```

```{r}
# chunk 15: you can also use the LabelClusters function to help label individual clusters
plot_seurat_clusters <- UMAPPlot(object = obj, label = TRUE, label.size = 5) +
                          labs(title = "Initial Clusters") +
                          theme(plot.title = element_text(hjust = 0.5),
                                axis.text = element_blank(),
                                axis.ticks = element_blank())
plot_seurat_clusters
```

```{r}
# chunk 16: some MANUAL work - sort by Cell_Type and Marker if not already sorted
markers_all <- read.csv(file = file.path(path_data, "marker_genes.csv"))
markers_all %>% slice_sample(n = 10) # example markers
```

```{r}
# chunk 17: 
select_cell_types <- c("B", "Macrophages", "mDC", "NK", "T")
select_markers <- c("CCR7", "CD19", "CD22", "CD31", "COL1A2", "DCN", "KRT1", "KRT5", "CD14", "MS4A7", "LYZ", "PMEL", "GNLY", "GZMA", "IRF7", "IRF8", "JCHAIN", "CD3D", "CD4", "CD8A", "IL7R" )
# do features = unique(markers_all$Marker) to use all possible features
p <- DotPlot(object = obj,
             features = markers_all %>%
                          dplyr::filter(Cell_Type %in% select_cell_types) %>%
                          #dplyr::filter(Marker %in% select_markers) %>%
                          distinct(Marker) %>% pull(),
             cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE)

# add in the cell type information
# if desired, you could rename the "Cell_Type"s in the original database to be
# more informative e.g. Natural_Killers instead of NK
p$data <- left_join(p$data,
                    markers_all %>%
                      dplyr::filter(Cell_Type %in% select_cell_types) %>%
                      dplyr::rename(features.plot = "Marker"),
                    by = "features.plot", multiple = "all")


# plot
p +
  facet_grid(cols = vars(Cell_Type), scales = "free_x", space = "free") +
  theme(strip.text.x = element_text(size = 10)) + RotatedAxis()
```

```{r}
# chunk 18: make a dot plot for clusters using ADT data
# you have to change the default assay for the dot plot
DefaultAssay(object = obj) <- "ADT"

DotPlot(object = obj,
        features = rownames(GetAssayData(object = obj, assay = "ADT",
                                         slot = "counts")),
        cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE) +
  RotatedAxis()
```

```{r}
# chunk 19: reset the default assay back to RNA!!
DefaultAssay(object = obj) <- "RNA"
```

```{r}
# chunk 20: feature plots 
FeaturePlot(object = obj,
            features = c("MS4A1", "CD14", # MS4A1 is another name for CD20
                         "NKG7", "IL3RA", "CD3D", "IL7R"), 
            min.cutoff = 0, label = TRUE, label.size = 3,
            ncol = 3, raster = FALSE)
```

```{r}
# chunk 21: violin plots
VlnPlot(object = obj,
        features = c("CD3-TotalSeqB", "CD14-TotalSeqB", "CD20-TotalSeqB",
                     "CD335-TotalSeqB"),
        pt.size = 0, assay = "ADT", ncol = 2) &
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title.x = element_blank())
```

```{r}
# chunk 22: name clusters 
# mDC = myeloid dendritic cells
# pDC = plasmacytoid dendritic cells
obj_annotations <- rbind(c("0", "Macrophages/mDCs"), # difficult to separate
                         c("1", "T Cells"),
                         c("2", "T Cells"),
                         c("3", "T Cells"),
                         c("4", "Natural Killers"),
                         c("5", "B Cells"),
                         c("6", "Macrophages/mDCs"), # difficult to separate
                         c("7", "Macrophages/mDCs"), # difficult to separate
                         c("8", "T Cells"), # faint signal
                         c("9", "pDCs"))
colnames(obj_annotations) <- c("Cluster", "CellType")
obj_annotations <- data.frame(obj_annotations)
obj_annotations
# save the annotations as a csv if you'd like
# write.csv(obj_annotations, file = file.path(path_data, "obj_annotations.csv"))

# prepare the annotation information
annotations <- setNames(obj_annotations$CellType, obj_annotations$Cluster)
annotations
# relabel the Seurat clusters
# Idents(obj) <- "seurat_clusters"
obj <- RenameIdents(object = obj, annotations)

# alphabetize the cell types
Idents(obj) <- factor(Idents(object = obj), levels = sort(levels(obj)))

# useful metadata (e.g. if you want to have multiple annotation sets)
obj[["annotated_clusters"]] <- Idents(object = obj)

# save the processed and annotated Seurat object
# saveRDS(obj, file = file.path(path_data, "tenx_pbmc5k_CITEseq_annotated.rds"))

# info about the clusters
obj_annotations %>%
  group_by(CellType) %>%
  transmute(Clusters = paste0(Cluster, collapse = ", ")) %>%
  distinct() %>%
  arrange(CellType)
```

```{r}
# chunk 23: overlay ADT information on top of feature plots to find B cells

FeaturePlot(object = obj, features = "adt_CD19-TotalSeqB",
            label = TRUE, label.size = 3, ncol = 1, raster = FALSE)
```
