---
title: "Single Cell Analysis"
author: "Max Mattessich, Joaquin Reyna, Edel Aron, Anna Konstorum"
date: "Compiled: `r format(Sys.time(), '%d %B %Y')`"
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amsfonts}
output:
    rmdformats::readthedown:
      code_folding: show
      dev: 'jpeg'
      df_print: kable
      fig_retina: 1
      highlight: kate
      toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Single-Cell-Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this vignette, we will cover some of the possible sources for
publicly available single-cell data, how to format it for and process it
with MCIA and how to run a basic analysis in order to annotate the data
and use the results as metadata for exploring the decomposition results.

![Vignette 2
Pipeline](https://github.com/Muunraker/nipalsMCIA/releases/download/v0.99.2/Vignette-2-Pipeline.png)

```{r load-packages, echo = TRUE, include = FALSE}
# note that the TENxPBMCData package is not included in this list as you may
# decide to pull data from another source or use our provided objects

library(dplyr)
library(ggplot2)
library(ggpubr)
library(piggyback)
library(Seurat)
```

```{r set-paths}
# recommended location for local data
path_data <- file.path("..", "data")
```

## Data

We will be using the 10x Genomics "pbmc5k-CITEseq" dataset, which was
published in May 2019 and processed using Cell Ranger 3.0.2. It contains
5,247 detected cells from PBMCs and 33,538 genes along with 32 cell
surface markers. We chose this dataset due to it containing both gene
expression (GEX) and cell surface protein (ADT) data as well as being
relatively recent, publicly available and from a widely used platform.

The following code details several ways in which to load in this
dataset. You may prefer to use data from other sources outside of 10x
Genomics, in which case you will have to format it to work with
nipalsMCIA.

### All Sources

We have the objects described below available in data files which you
can download and load in here if you do not want to run the following
sections. **These files includes the processed object that you will need
in order to run MCIA (data_blocks_sc.Rda).**

```{r data-all-download, eval = FALSE}
# marker genes for cell type annotation with Seurat
pb_download(file = "marker_genes.csv", dest = path_data,
            repo = "Muunraker/nipalsMCIA")

# the Seurat data's metric summary file from 10x Genomics
pb_download(file = "5k_pbmc_protein_v3_metrics_summary.csv", dest = path_data,
            repo = "Muunraker/nipalsMCIA")

# Seurat objects in different stages of processing
pb_download(file = "tenx_pbmc5k_CITEseq_raw.rds", dest = path_data,
            repo = "Muunraker/nipalsMCIA")
pb_download(file = "tenx_pbmc5k_CITEseq_annotated.rds", dest = path_data,
            repo = "Muunraker/nipalsMCIA")
```

### 10x Genomics and Seurat

The original dataset can be found on the [10x Genomics Datasets
website](https://support.10xgenomics.com/single-cell-gene-expression/datasets/3.0.2/5k_pbmc_protein_v3)
and an explanation of the file types for this version can be found
[here](https://support.10xgenomics.com/single-cell-gene-expression/software/pipelines/3.0/output/overview).
You can download them all to a directory of your choosing with their
suggested terminal commands:

```{r data-10x-download, eval = FALSE}

# download the raw feature matrix
url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_raw_feature_bc_matrix.tar.gz"
dest = file.path(path_data, "5k_pbmc_protein_v3_raw_feature_bc_matrix.tar.gz")
download.file(url,      destfile = dest, method = "auto")

# download the filtered feature matrix
url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz"
dest = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz")
download.file(url, destfile = dest, method = "auto")


# download the metrics summary
url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_metrics_summary.csv"
dest = file.path(path_data, "5k_pbmc_protein_v3_metrics_summary.csv")
download.file(url, destfile = dest, method = "auto")


# download the summary html (optional)
url = "https://cf.10xgenomics.com/samples/cell-exp/3.0.2/5k_pbmc_protein_v3/5k_pbmc_protein_v3_web_summary.html"
dest = file.path(path_data, "5k_pbmc_protein_v3_web_summary.html")
download.file(url, destfile = dest, method = "auto")
```

If you would like to view some basic information about the data, you can
open the *web_summary.html* in your favorite web browser. It will show
the estimated number of cells, mean reads per cell, median genes per
cell and a variety of other sample metrics. These metrics are also
available in the *metrics_summary.csv* file.

You will need the *filtered_feature_bc_matrix* directory for the
following analysis; it can be extracted with
`tar -xvzf 5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz` from
within the relevant data directory.

```{r}

tar_fn = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix.tar.gz")
#exdir = file.path(path_data, "5k_pbmc_protein_v3_filtered_feature_bc_matrix")
exdir = file.path(path_data, "tenx_pbmc5k_CITEseq")
untar(tar_fn, exdir = exdir, tar = "internal")
```

```{r data-10x-object, eval = TRUE}
# load the data (change the file path as needed)
data <- Seurat::Read10X(data.dir = file.path(path_data, "tenx_pbmc5k_CITEseq",
                                             "filtered_feature_bc_matrix"),
                        strip.suffix = TRUE) # remove the "-1"s from barcodes
## 10X data contains more than one type and is being returned as a list
## containing matrices of each type.

# set minimum cells and/or features here if you'd like
obj <- Seurat::CreateSeuratObject(counts = data$`Gene Expression`,
                                  project = "pbmc5k_CITEseq")
obj[["ADT"]] <- Seurat::CreateAssayObject(counts = data$`Antibody Capture`)
## Warning: Feature names cannot have underscores ('_'), replacing with dashes ('-')

# check the assays
Seurat::Assays(object = obj)
## "RNA" "ADT"

# list out the CITE-Seq surface protein markers
rownames(obj[["ADT"]])
```

You can also load the Seurat object directly, such as in the [Read in
and process the data] subsection in the [Deep dive: Seurat analysis]
section later in the vignette.

## Deep dive: Seurat analysis

This section demonstrates how to take in raw data (in this case, the
output of Cell Ranger v3.0) and go through a popular analysis pipeline
to ultimately cluster and annotate the data. Some people prefer to use
Cell Ranger's built-in dimension reduction and clustering analysis and
to view the results with [Loupe Cell
Browser](https://support.10xgenomics.com/single-cell-gene-expression/software/visualization/latest/what-is-loupe-cell-browser).

*Credit to Seurat
([RNA](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html) and
[multi-modal](https://satijalab.org/seurat/articles/multimodal_vignette.html#setup-a-seurat-object-add-the-rna-and-protein-data))
for the general steps.*

### Read in and process the data

```{r seurat-load-data, message = FALSE}
# load the data

# add useful metadata
obj[["percent.mt"]] <- PercentageFeatureSet(object = obj, pattern = "^MT-")
```

### Quality control

#### Metrics summary

```{r seurat-qc-metrics-summary}
# read in the summary table
(metrics_summary <- read.csv(
    file = file.path(path_data, "5k_pbmc_protein_v3_metrics_summary.csv")
))
```

#### GEX QC metrics

##### Before filtering

```{r seurat-qc-pre-filter, fig.dim = c(8, 5)}
VlnPlot(object = obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        pt.size = 0.01, ncol = 3) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title.x = element_blank())
```

##### After filtering

Based on the previous plots, a minimum of 200 features and a maximum of
20% mitochondrial seemed like good cutoffs.

```{r seurat-qc-post-filter, fig.dim = c(8, 5)}
# adjust cutoffs as desired
obj <- subset(x = obj, subset = nFeature_RNA > 200 & percent.mt < 20)

VlnPlot(object = obj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        pt.size = 0.01, ncol = 3) &
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title.x = element_blank())
```

### Standard Seurat pipeline

**Most of these are run on the RNA**.

```{r seurat-pipeline-1, eval = TRUE}
# standard log normalization for RNA and centered log for ADT
obj <- NormalizeData(object = obj, normalization.method = "LogNormalize",
                     scale.factor = 10000, assay = "RNA")
```

```{r}
obj <- NormalizeData(object = obj, normalization.method = "CLR",
                     margin = 2, assay = "ADT") # go across cells, not 
```

```{r}

# highly variable features
obj <- FindVariableFeatures(object = obj, selection.method = "vst",
                            nfeatures = 2000)
```

```{r}
# scaling so the average expression is 0 and the variance is 1
obj <- ScaleData(object = obj, features = rownames(obj))
```

```{r}
# dimensionality reduction
obj <- RunPCA(object = obj)
```

```{r}
# clustering (adjust dimensions and resolutions as desired)
obj <- FindNeighbors(object = obj, reduction = "pca", dims = 1:20)
obj <- FindClusters(object = obj, resolution = 0.4)
```

```{r}
obj <- RunUMAP(object = obj, reduction = "pca", dims = 1:20)
```

### Dimensionality reduction

#### Load in the processed object

To avoid having to save multiple large objects, this file already
includes the annotations defined in a later section, so we will clear
them out here to proceed as normal.

```{r seurat-load-processed-object}
# reset to baseline
Idents(object = obj) <- "seurat_clusters"
#obj$annotated_clusters <- c()
```

#### PCA

**Typically this section would go in between the `RunPCA()` and
`FindNeighbors()` steps above**. It has been included here because of
how we save and load the objects separately for processing speed.

Now that we've run PCA, we can examine an elbow plot as a simple method
for selecting how many PCs to choose when identifying
neighbors/clusters.

```{r seurat-elbow, fig.dim = c(6, 3),fig.align='center'}
ElbowPlot(object = obj, ndims = 30)
```

From the plot, it looked like 20 works as a cut-off for the number of
PCs to include. We could have also chosen to use 10, but decided to go a
little higher.

You can also examine other information which was used for the PCA, such
as the top twenty most variable features:

```{r seurat-top-features}
head(VariableFeatures(object = obj), 20)
```

```{r seurat-top-features-plot, fig.dim = c(6, 6), warning = FALSE, eval = TRUE}
# these can be plotted if desired
LabelPoints(plot = VariableFeaturePlot(object = obj),
            points = head(VariableFeatures(object = obj), 20),
            repel = TRUE, xnudge = 0, ynudge = 0) +
  labs(title = "Top Twenty Variable Features") +
  theme(legend.position = "bottom")
```

#### UMAP

```{r seurat-umap-initial, fig.dim = c(8, 8),fig.align='center'}
# you can also use the LabelClusters function to help label individual clusters
plot_seurat_clusters <- UMAPPlot(object = obj, label = TRUE, label.size = 5) +
                          labs(title = "Initial Clusters") +
                          theme(plot.title = element_text(hjust = 0.5),
                                axis.text = element_blank(),
                                axis.ticks = element_blank())
plot_seurat_clusters
```

### Marker overlays

Note that the following marker genes **are not meant to be an exhaustive
list.** They are included as examples of some of the cell types you
could look for.

#### Load marker genes

```{r seurat-markers-load}
# sort by Cell_Type and Marker if not already sorted
markers_all <- read.csv(file = file.path(path_data, "marker_genes.csv"))
markers_all %>% slice_sample(n = 10) # example markers
```

#### Dot plots

##### GEX

Here we only demonstrate a few of the cell types included within the
provided markers database. We plot those cell types on top of the dot
plot to make it easier to see which markers they correspond with, but
you can comment that code out if you would rather have a simple dot
plot.

```{r seurat-markers-dot-gex, fig.dim = c(14, 6), message = FALSE}
select_cell_types <- c("B", "Macrophages", "mDC", "NK", "T")

# do features = unique(markers_all$Marker) to use all possible features
p <- DotPlot(object = obj,
             features = markers_all %>%
                          dplyr::filter(Cell_Type %in% select_cell_types) %>%
                          distinct(Marker) %>% pull(),
             cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE)

# add in the cell type information
# if desired, you could rename the "Cell_Type"s in the original database to be
# more informative e.g. Natural_Killers instead of NK
p$data <- left_join(p$data,
                    markers_all %>%
                      dplyr::filter(Cell_Type %in% select_cell_types) %>%
                      dplyr::rename(features.plot = "Marker"),
                    by = "features.plot", multiple = "all")

# plot
p +
  facet_grid(cols = vars(Cell_Type), scales = "free_x", space = "free") +
  theme(strip.text.x = element_text(size = 10)) + RotatedAxis()
```

##### ADT

The average expression was set to a minimum of zero to better see the
up-regulated features.

```{r seurat-markers-dot-adt, fig.dim = c(14, 6)}
# you have to change the default assay for the dot plot
DefaultAssay(object = obj) <- "ADT"

DotPlot(object = obj,
        features = rownames(GetAssayData(object = obj, assay = "ADT",
                                         slot = "counts")),
        cols = "RdBu", col.min = -1, dot.scale = 3, cluster.idents = TRUE) +
  RotatedAxis()

# reset the default
DefaultAssay(object = obj) <- "RNA"
```

#### Feature plots

If you would like to generate feature plots for a cell type within your
markers database (if you are using one) e.g. for B cells, you could do
`features = (dplyr::filter(markers_all, Cell_Type == "B"))$Marker` in
the following command. Here we just show a few characteristic markers
for different cell types for simplicity.

```{r seurat-markers-feature, fig.dim = c(12, 8)}
FeaturePlot(object = obj,
            features = c("MS4A1", "CD14", # MS4A1 is another name for CD20
                         "NKG7", "IL3RA", "CD3D", "IL7R"), 
            min.cutoff = 0, label = TRUE, label.size = 3,
            ncol = 3, raster = FALSE)
```

#### Violin plots

Here the point size is set to zero so that you can just see the violins.
We only show a few of the cell surface markers here, but you could plot
them all with `features = rownames(obj[["ADT"]])`. You can also plot the
standard GEX markers just like for the feature plots (just make sure to
remove `assay = "ADT"`).

```{r seurat-markers-violin-adt, fig.dim = c(8, 6)}
VlnPlot(object = obj,
        features = c("CD3-TotalSeqB", "CD14-TotalSeqB", "CD20-TotalSeqB",
                     "CD335-TotalSeqB"),
        pt.size = 0, assay = "ADT", ncol = 2) &
  theme(plot.title = element_text(size = 10),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        axis.title.x = element_blank())
```

### Annotate cell clusters

These annotations are chosen here more as a proof of general concept
instead of a more highly refined and verified approach that you would
typically see within a single-cell publication. There are also
[automated methods for
annotation](https://www.sciencedirect.com/science/article/pii/S2001037021000192).

#### Annotations

If you would like to further break down these clusters, you can increase
the clustering resolution in the standard pipeline in order to increase
the number of clusters.

```{r seurat-annotations}
# mDC = myeloid dendritic cells
# pDC = plasmacytoid dendritic cells
obj_annotations <- rbind(c("0", "Macrophages/mDCs"), # difficult to separate
                         c("1", "T Cells"),
                         c("2", "T Cells"),
                         c("3", "T Cells"),
                         c("4", "Natural Killers"),
                         c("5", "B Cells"),
                         c("6", "Macrophages/mDCs"), # difficult to separate
                         c("7", "Macrophages/mDCs"), # difficult to separate
                         c("8", "T Cells"), # faint signal
                         c("9", "pDCs"))
colnames(obj_annotations) <- c("Cluster", "CellType")
obj_annotations <- data.frame(obj_annotations)

# save the annotations as a csv if you'd like
# write.csv(obj_annotations, file = file.path(path_data, "obj_annotations.csv"))

# prepare the annotation information
annotations <- setNames(obj_annotations$CellType, obj_annotations$Cluster)

# relabel the Seurat clusters
# Idents(obj) <- "seurat_clusters"
obj <- RenameIdents(object = obj, annotations)

# alphabetize the cell types
Idents(obj) <- factor(Idents(object = obj), levels = sort(levels(obj)))

# useful metadata (e.g. if you want to have multiple annotation sets)
obj[["annotated_clusters"]] <- Idents(object = obj)

# save the processed and annotated Seurat object
# saveRDS(obj, file = file.path(path_data, "tenx_pbmc5k_CITEseq_annotated.rds"))

# info about the clusters
obj_annotations %>%
  group_by(CellType) %>%
  transmute(Clusters = paste0(Cluster, collapse = ", ")) %>%
  distinct() %>%
  arrange(CellType)

# metadata file for MCIA
meta <- data.frame("CellType" = Idents(object = obj))

# save the metadata file for MCIA
# write.csv(meta, file = file.path(path_data, "metadata_sc.csv"))
```

#### UMAPs

Here we use the selected metadata colors from the [MCIA] section, so be
sure to change that (or just remove the `cols = ...` to use the default)
if you are running this section first.

```{r seurat-umap-both, fig.dim = c(16, 8)}
# change colors and themes as desired
# plot_annotated_clusters <- UMAPPlot(object = obj,
#                                     label = TRUE, label.size = 4,
#                                     cols = meta_colors_sc) +
#                              labs(title = "Annotated Clusters") +
#                              theme(plot.title = element_text(hjust = 0.5),
#                                    axis.text = element_blank(),
#                                    axis.ticks = element_blank())
# 
# ggarrange(plot_seurat_clusters, plot_annotated_clusters)
```

You should change the `DefaultAssay` to/from `ADT` like in [ADT] if you
don't want to specify "adt\_" before the feature name.

```{r seurat-umap-adt, fig.dim = c(4, 4)}
# plot ADT information on top as an example

FeaturePlot(object = obj, features = "adt_CD19-TotalSeqB",
            label = TRUE, label.size = 3, ncol = 1, raster = FALSE)
```

#### Check the annotations

For example, if you want to visually check your T cell cluster
annotations (just like before, you can also do
`features = (dplyr::filter(markers_all, Cell_Type == "T"))$Marker` to
use all of the T cell markers in the markers database):

```{r seurat-check-annotations, fig.dim = c(8, 6)}
# VlnPlot(object = obj,
#         features = c("CD3D", "CD4", "IL7R", "TRAC"),
#         cols = meta_colors_sc, pt.size = 0.01, ncol = 2) &
#   theme(plot.title = element_text(size = 10),
#         axis.title.x = element_blank())
```

## Session Info

```{r session-info}
sessionInfo()
```
