---
title: "Week 2 - Class 0 - SLE mini project"
format: html
editor: visual
---

## SLE mini project (continued)

Repeating some of the previous class (week 1)

If necessary use:

```         
install.packages('BiocManager')
```

then `BiocManager::install("GEOquery")` function call to install the package. We will also use **dplyr** and **ggplot2** packages to help explore this dataset further.

NOTE: **select** function confuses R sometimes. When you refer to the **select** within **dplyr** it is a good idea to use **dplyr::select** instead of **select**

```{r}
# chunk 1: read, clean, filter the data for a specific cell type

# Load the required packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("DESeq2")
BiocManager::install("GEOquery")
# Load the required packages
library(GEOquery)
library(dplyr)
library(ggplot2)
library(DESeq2)

path_data <- file.path("..", "data")
path_output <- file.path("..", "output")
dir.create(path_output, showWarnings = FALSE)

# Complete the code with the GEO ID
gse <- getGEO("GSE149050")
gse

# Our gse list has only one entry
metadata <- pData(phenoData(gse[[1]]))
dim(metadata)

grep("characteristics", colnames(metadata), value=TRUE)

table(metadata$characteristics_ch1)[1]

table(metadata$characteristics_ch1.2, metadata$characteristics_ch1.7)

#metadata.cDC <- filter(metadata, characteristics_ch1.2 == "cell type: cDC" &
#                       characteristics_ch1.6 == "visit number: 1")
#head(metadata.cDC[,1:3])

metadata.cmo <- filter(metadata, characteristics_ch1.2 == "cell type: cMo" &                        characteristics_ch1.6 == "visit number: 1")
head(metadata.cmo[,1:3])

# TODO: Build up this example by piping to head() after each line
# Cleans up headers to our sample names
metadata.subset <- metadata.cmo %>%
  dplyr::select(title,
         disease_state = characteristics_ch1, 
         ifn_status = characteristics_ch1.3,
         patient_id = characteristics_ch1.4) %>%
  mutate(disease_state = gsub("disease state: ","", disease_state)) %>%
  mutate(ifn_status = gsub("ifn status: ","", ifn_status)) %>%
  mutate(patient_id = gsub("patientuid: ","", patient_id)) %>%
  mutate(state = recode(disease_state, 
                        "healthy control" = "control", 
                        "systemic lupus erythematosus (SLE)" = "lupus") )
						
table(metadata.subset$state)

counts.all <- read.delim(file.path(path_data, "GSE149050_Bulk_Human_RawCounts.txt"), check.names=FALSE, row.names = 1)

head(counts.all[,1:3])
# below is one of the data objects that will go into DESeq
counts.subset <- counts.all %>% dplyr::select(metadata.subset$title) 
head(counts.subset[,1:4])
```

## DESeq

For DESeq we need to do some book-keeping and get the data in the correct format. First double check on column correspondences between counts and metadata

```{r}
# chunk 2: sanity check
all(colnames(counts.subset) == metadata.subset$title)
#all(c(1,2,3,4) == c(1,2,3,4))
```

Finally, DESeq needs our sample names as rownames of the metadata.

```{r}
# chunk 3: shuffle columns around so DESeq is happy
# Remove the title column then use it as rownames
colData <- metadata.subset[,-1] # drops first column
rownames(colData) <- metadata.subset[,1] # keeping only that column
head(colData)
```

Now we can setup our DESeq object with the long but appropriately named `DESeqDataSetFromMatrix()` function:

```{r}
# chunk 3: shuffle columns around so DESeq is happy
#TODO: look up how to use DESeq design argument
dds <- DESeqDataSetFromMatrix(countData = counts.subset,
                              colData = colData,
                              design = ~state)
```

Before we go any further we should remove any genes that have very low read counts. Here we will follow general practice of keeping rows that have at least 10 reads total. This will speed up our future calculations and avoid diluting our statistical power by doing needless tests on genes that we can not say much about anyway.

```{r}
# chunk 4: remove genes with very low counts
# these would have never come up as statistically significant
# BUT they impact the number of tests conducted hence the adj. pval
#Ferhat changed this to 100, more stringent, smaller dataset
keep.inds <- rowSums(counts(dds)) >= 10
dds <- dds[keep.inds,]
```

Now let's run the DESeq analysis pipeline on the dataset, and reassign the resulting object back to the same `dds` object:

```{r}
# chunk 5: All-in-one - running DESeq the main function!
dds <- DESeq(dds)
# equivalent to below (AKA older functions):
#dds <- estimateSizeFactors(dds)
#dds <- estimateDispersions(dds)
#dds <- nbinomWaldTest(dds)

```

Here, we're running the DESeq pipeline on the `dds` object and it can take some time. The `DESeq()` function calls a number of other functions within the package to essentially run the entire pipeline (normalizing by library size by estimating the "size factors," estimating dispersion for the negative binomial model, and fitting models and getting statistics for each gene for the design specified when we imported the data).

DESeq2 offers two kinds of hypothesis tests: the Wald test, where we use the estimated standard error of a log2 fold change to test if it is equal to zero, and the likelihood ratio test (LRT). The LRT examines two models for the counts, a *full* model with a certain number of terms and a *reduced* model, in which some of the terms of the *full* model are removed. The test determines if the increased likelihood of the data using the extra terms in the *full* model is more than expected if those extra terms are truly zero

Since we've got a fairly simple design (single factor, two groups, lupus vs control), we can get results out of the object simply by calling the `results()` function on the DESeqDataSet that has been run through the pipeline.

The help page for `?results` and the vignette both have extensive documentation about how to pull out the results for more complicated models (multi-factor experiments, specific contrasts, interaction terms, time courses, etc.).

```{r}
# chunk 6: take a peek at all the results - top of the data structure only
res <- results(dds) 
head(res)
```

We can summarize some basic tallies using the `summary()` function. And with a different `alpha` (adjusted p-value) threshold of 0.05:

```{r}
# chunk 7: summarize before and after filtering for significant DE genes
summary(res)
res_p05 <- results(dds, alpha=0.05)
summary(res_p05)
```

Let's finally extract our top genes of interest!

```{r}
# chunk 8: filter genes more and create DESeq output files 
# only extracting "upregulated genes" in SLE and at least 2 fold change
top.genes <- as.data.frame(res) %>% filter(padj <= 0.05 & log2FoldChange >= 1)
dim(top.genes)
write.csv(top.genes, file=file.path(path_output, "deseq_results_topgenes.csv"))

#save(top.genes, file = "top_genes.RData")

bot.genes <- as.data.frame(res) %>% filter(padj <= 0.05 & log2FoldChange <= -1)
dim(bot.genes)
write.csv(bot.genes, file=file.path(path_output, "deseq_results_botgenes.csv"))


# Comprehensive file with all genes and DE statistics
# TO BE USED in the next exercise
write.csv(res, file=file.path(path_output, "deseq_results.csv"))


```

```{r}
results <- as.data.frame(res)
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 1] = "blue"
mycols[ abs(res$log2FoldChange) < -1] = "red"

mycols[ res$padj > 0.05 ] = "gray"

ggplot(results) +
  aes(log2FoldChange, -log(padj)) +
  geom_point(col=mycols) +
  geom_vline(xintercept = c(-1,+1), linetype=2) +
  geom_hline(yintercept = -log(0.05), linetype=2)
```

```         
```

OPTIONAL: compare IFN positive vs IFN negative samples

```{r}

table(colData$ifn_status)
# Select only IFNneg and IFNpos count columns in colData
ifn_colData <- filter(colData, ifn_status %in% c("IFNneg", "IFNpos"))
#ifn_colData <- filter(colData, ifn_status %in% c("HC", "IFNpos"))
table(ifn_colData$ifn_status)

# Select only IFNneg and IFNpos count columns in counts data
ifn_countData <- dplyr::select(counts.subset, rownames(ifn_colData))
dim(ifn_countData)

dim(ifn_colData)

# create DEseq dds object and run DESeq
ifn_dds <- DESeqDataSetFromMatrix(countData = ifn_countData,
                              colData = ifn_colData,
                              design = ~ifn_status)

ifn_dds <- DESeq(ifn_dds)
ifn_res <- results(ifn_dds)

```

```{r}
summary(ifn_res)
ifn_res_p05 <- results(ifn_dds, alpha=0.05)
summary(ifn_res_p05)
```

```{r}
top.genes <- as.data.frame(ifn_res) %>% filter(padj <= 0.05 & log2FoldChange >= 1)
dim(top.genes)
write.csv(top.genes, file=file.path(path_output, "ifn_deseq_results_topgenes.csv"))

write.csv(ifn_res, file=file.path(path_output, "ifn_deseq_results.csv"))
```
